---
title: 'Food for Thought: An Analysis of Everyone''s Favorite Morning Nectar, Coffee'
author: "Ashrita Kodali"
date: "2023-11-24"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=12, fig.height=8)
knitr::opts_knit$set(root.dir = "/Users/ashik/1st Year UVA/STAT 3280")
```

```{r, echo=FALSE,  warning=FALSE, message=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(plotly)
library(patchwork)
library(ggpie)
library(webr)
```

# Introduction:  

C8H10N4O2. Caffeine is found in many beverages including energy drinks, soda’s, and America’s second favorite drink: coffee. Coffee is the morning nectar that the United States runs on and is a significant part of my life. More often than not, I can be found at cafes studying and sipping my coffee throughout the day. My habits made me curious on how much coffee I consume day to day and  how I compare to the average adult coffee drinker in the US. My goal for this project is to displace a narrative that tracks not only my own coffee habits but compares them to my broader population of coffee drinkers in the United States. As such, I found a dataset describing my population. The data for the population comes from a Youtuber who gathered responses regarding the coffee habits, consumption, and other auxiliary variables for  nearly 4000 individuals in order to conduct their experiment. I used the auxiliary variables the youtuber collected in order to create comparisons across the data I collected for myself. It is important to note that most of the responses and questions for the data on the population were phrased in such a way that categorical data was collected. As such, many pie charts and bar graphs were created in order to accurately represent the data. 

```{r, echo=FALSE,  warning=FALSE, message=FALSE}
coffee_pop <- read.csv("coffee population.csv")
coffee_pop$How.many.cups.of.coffee.do.you.typically.drink.per.day. <- 
  recode_factor(
    coffee_pop$How.many.cups.of.coffee.do.you.typically.drink.per.day.,
    "Less than 1" = "0", "More than 4" = "5"
  )
coffee_pop$How.many.cups.of.coffee.do.you.typically.drink.per.day. <-
  as.numeric(coffee_pop$How.many.cups.of.coffee.do.you.typically.drink.per.day.)

coffee_me <- read.csv("coffee_data_personal.csv")
coffee_me$date. <- as.Date(coffee_me$date., format="%m/%d/%Y")
```
### Comparisons Between Myself and the Population: 

##### Figure 1 -  America's Favorite Drink:  

```{r, echo=FALSE}
color <- c("slategray4", "slategray4", "slategray4", "slategray4", "slategray4",
           "slategray4", "slategray4", "lightpink", "slategray4", "slategray4",
           "lightpink", "slategray4")
data <- coffee_pop[coffee_pop$What.is.your.favorite.coffee.drink. != "",]
ggplot(data, aes(x=What.is.your.favorite.coffee.drink., 
                 fill=What.is.your.favorite.coffee.drink.)) + 
  geom_bar(color="black") +
  xlab("Coffee Drinks") + ylab("Count") + 
  ggtitle("Bar Chart of Adult Americans' Favorite Coffee Drink by Counts") +
  theme(axis.text.x = element_text(angle=38, size=10, color="black"), 
        legend.position = "none") +
  scale_fill_manual(values=c("slategray4", "slategray4", "slategray4",
                             "slategray4", "slategray4", "slategray4",
                             "slategray4", "lightpink", "slategray4",
                             "slategray4", "lightpink", "slategray4")) +
  scale_y_continuous(breaks=seq(from=0, to=1000, by=100))
```
*Figure 1 is a bar chart of the most popular drinks beloved by adult coffee drinkers in the United States. In the bar chart, the pour over followed by the latte (both of which are highlighted in pink) are the most popular drinks by those who drink coffee. *

##### Figure 2 - My Favorite Drink: 

```{r, echo=FALSE}
ggplot(coffee_me, aes(x=order, fill=order)) + geom_bar(color="black") + 
  xlab("Coffee Drinks") + ylab("Count") + 
  ggtitle("Bar Chart of My Favorite Coffee Drink by Counts") +
  scale_fill_manual(values=c("slategray4","slategray4", "lightblue",
                             "slategray4", "slategray4", "slategray4",
                             "slategray4", "slategray4", "lightblue")) + 
  theme(axis.text.x = element_text(angle=30, size=10, color="black"), 
        legend.position = "none") + 
  scale_y_continuous(breaks=seq(0,15,1))
```

*Figure 2 is a bar chart of the number of times I made or bought a coffee drink. In the bar chart, any version of cold brew (the vanilla sweet cream cold brew and cold brew with creamer, both of which are highlighted in light blue) are my favorite drinks because I bought or made them the most frequently.*

When comparing myself with the average adult coffee drinker in the US, my favorite beverage does not align with the consensus of those in the US. I seem to prefer drinking cold brew coffee while the average drinker in the United States seems to enjoy the occasional pour over or a latte. Although the latte was ranked the second favorite, many studies, including that from the National Coffee Association of America, stated that lattes are actually the most popular coffee drink. In either case, I seemingly do not align with the average coffee drinker, setting me apart from the population. 

##### Figure 3 - Where Is Everyone Drinking Their Coffee? :  

```{r, echo=FALSE}
home1 <- coffee_pop %>% group_by(What.is.your.age.) %>%
  summarise(home = sum(Where.do.you.typically.drink.coffee...At.home., na.rm=TRUE),
            office = sum(Where.do.you.typically.drink.coffee...At.the.office., na.rm = TRUE),
            go = sum(Where.do.you.typically.drink.coffee...On.the.go., na.rm = TRUE),
            cafe = sum(Where.do.you.typically.drink.coffee...At.a.cafe., na.rm = TRUE),
            avg_cups = sum(How.many.cups.of.coffee.do.you.typically.drink.per.day.)/n())
sum_places <- c(home1$home, home1$office, home1$go, home1$cafe)
age_var <- c(home1$What.is.your.age., home1$What.is.your.age.,
             home1$What.is.your.age.,home1$What.is.your.age.)
places <- c("home", "home", "home", "home", "home", "home", "home", "home",
            "office", "office", "office", "office", "office", "office", "office", "office",
            "go", "go", "go", "go", "go", "go", "go", "go",
            "cafe", "cafe", "cafe", "cafe", "cafe", "cafe", "cafe", "cafe")
df <- data.frame(age_var, places, sum_places)
df <- df[df$age_var != "",]
ggplot(df, aes(y=sum_places, x=places, fill=places)) + geom_bar(position="dodge",
                                                                 stat="identity") +
  facet_grid(cols=vars(age_var)) + facet_wrap(vars(age_var), nrow = 2) +
  xlab("Places Where People Drank Coffee") + 
  ylab("Total Number of Votes for Each Place") +
  ggtitle("Bar Chart of Total Votes for Each Location Adults in the United States Drank Their Coffee by Each Age Group")
```

*Figure 3 is a bar chart of where the average adult coffee drinker in the United States drinks their coffee categorized by each age group. In all of the age groups, most coffee drinkers tend to drink their coffee at home*

##### Figure 4 - Where Am I Drinking Coffee? : 
```{r, echo=FALSE}
places_me <- coffee_me %>% group_by(where) %>%
  summarise(counts = n())
ggplot(places_me, aes(x="", y=counts, fill=where)) + geom_bar(stat="identity") +
  coord_polar("y", start=0) + scale_fill_discrete() + theme_void() +
  geom_text(aes(x = 1, label = where), size=8, color = "white", 
            position = position_stack(vjust = .5)) +
  ggtitle("Pie Chart of Where I Drink Coffee") +
  theme(title = element_text(size=20),
        legend.position = "none")
```

*Figure 4 is a pie chart of where I drink my coffee. In all, I tend to enjoy my coffee in a cafe*

When comparing myself with the average adult coffee drinker in the US, my favorite location to drink coffee does not align with the location where the average adult drinks their coffee. I seem to prefer drinking in a cafe-setting while most people actually drink their coffee at home. This indicates that the spaces where we like enjoying our beverage are vastly different because I am more likely to be in a cafe drinking coffee while most American coffee drinkers will be drinking their coffee at home. 

##### Figure 5 - How Many Cups of Joe is too Many Cups of Joe?
Shiny App Link: https://ashritakodali.shinyapps.io/coffeehabits/

*Figure 5 is an interactive visualization in which the density of the number of cups of coffee consumed can be displayed for the entire population or by each specific age range. In order to get more specific details, the mean number of cups of coffee consumed per age range and the associated standard deviation can also be displayed*

##### Figure 6 - Am I a Coffee Addict? :  

```{r, echo=FALSE}
ggplot(coffee_me, aes(x=cups)) + geom_density(fill="lightblue", alpha=0.7) +
  xlab("Number of Coffee Cups") + ylab("Density") +
  ggtitle("Density Plot Distribution of Number of Coffee Cups I Drank") + 
  scale_y_continuous(breaks=seq(0,1,0.10)) + 
  geom_text(x=1.32, y=1, label="16 fluid oz (4/3 cups)", family="Courier", fontface="italic") + 
  geom_text(x=1.955, y=0.91, label="24 fluid oz (2 cups)",family="Courier", fontface="italic")
```

*Figure 6 is also a density plot of the number of cups of coffee I consume. There are two key annotations that point to peaks of the plot that indicate my “favorite amount” of coffee to consume.*

When comparing myself to the average coffee drinker, I shockingly drink less coffee than the average American adult drinker. Looking at the population, the mean number of coffee drinks the average American adult drinks is approximately 4.325 cups. However, I only drink 1.5333 cups of coffee everyday on average. In fact, the National Coffee Association of America indicates that the average American “drinks over 3 cups of coffee” every day. While I expected to be drinking more than the average coffee drinker, I am significantly under the average and thus vastly vary from the population of American coffee drinkers in the amount of coffee I consume. 

##### Figure 7 - Why Are We All Drinking? : 

```{r, echo=FALSE,  warning=FALSE, message=FALSE}
df <- coffee_pop[coffee_pop$What.is.your.age. != "",]
reasons <- df %>% group_by(What.is.your.age.) %>%
  summarise(taste = sum(Why.do.you.drink.coffee...It.tastes.good., 
                        na.rm=TRUE),
            energy = sum(Why.do.you.drink.coffee...I.need.the.caffeine.,
                         na.rm=TRUE),
            ritual = sum(Why.do.you.drink.coffee...I.need.the.ritual., 
                         na.rm=TRUE),
            bathroom = sum(Why.do.you.drink.coffee...It.makes.me.go.to.the.bathroom.,
                           na.rm=TRUE),
            avg_cups = sum(How.many.cups.of.coffee.do.you.typically.drink.per.day.)/n())
reason <- c("taste", "taste", "taste", "taste", "taste", "taste", "taste",
            "energy", "energy", "energy", "energy", "energy", "energy", "energy",
            "ritual", "ritual", "ritual", "ritual", "ritual", "ritual", "ritual",
            "bathroom", "bathroom", "bathroom", "bathroom", "bathroom", 
            "bathroom", "bathroom")
reason_counts <- c(reasons$taste, reasons$energy, reasons$ritual, reasons$bathroom)
age.ranges <- c(reasons$What.is.your.age., reasons$What.is.your.age.,
             reasons$What.is.your.age., reasons$What.is.your.age.)
df <- data.frame(age.ranges, reason, reason_counts)
df <- df[(df$age.ranges == "35-44 years old" )| (df$age.ranges == "18-24 years old")| 
           (df$age.ranges == "25-34 years old"),]
grouped_df <- df %>% group_by(age.ranges, reason)
PieDonut(grouped_df, aes(age.ranges, reason, count = reason_counts), r0=0.5)
```

*Figure 7 is a nested pie chart that first divides into 3 age groups and then divides each age group by their reasoning for drinking coffee. The four reasons include taste, their daily routine/ritual, for energy, and to go to the bathroom. Of the 4 reasons, most adults seem to enjoy their coffee for the taste of it.*

##### Figure 8 - Why Am I Drinking Coffee? : 

```{r, echo=FALSE}
ggplot(coffee_me, aes(x=reasoning)) + geom_bar(position="dodge",
                                               fill = "lightblue", 
                                               color="black") + 
  xlab("Reasons For Getting Coffee") +
  ylab("Count") + ggtitle("Counts of Each Reason I Got Coffee") + 
  scale_y_continuous(breaks=c(0:15))
```

*Figure 8 is a bar chart with the counts for each reason indicating why I got coffee. My four reasons include coffee being my breakfast, coffee being my lunch, for the taste/desire of it, or because I needed energy. Of the 4 reasons, needing energy was my most frequent reason on why I drank coffee*

When comparing myself to the average coffee drinker, I drink coffee for different reasons than the average American adult drinker. Looking at the population, most coffee drinkers tend to drink coffee because they like the taste of it. However, I decide to drink coffee because I tend to use it for energy or to feel the caffeine rush so I can concentrate and finish my work. In all, the average US coffee drinker and I have different reasons and purposes for drinking coffee. 

### Fun Facts:

##### Figure 9 - Money Does In Fact Buy Happiness: 

```{r, echo=FALSE, warning=FALSE, message=FALSE}

animated_plot <- plot_ly(data=coffee_me, x=~cost, y=~happiness,
                       type="scatter", mode="marker",
                       marker = list(sizemode = "diameter"),
                       frame=~week) %>% 
  layout(title = 'Scatterplot of the Cost of Each Drink and my Associated Happiness',
           xaxis  = list(title="Cost of Coffee (in Dollars)"), 
           yaxis = list(title="Happiness (Scale from 1 to 5)"))
animated_chart <- animated_plot %>% 
  animation_opts(frame = 1000,
                 redraw = TRUE, 
                 easing = "linear")
interactive_chart <- animated_chart %>% layout(
  updatemenus = list(
    list(
      type = "buttons",
      showactive = FALSE,
      x = 0,
      y = -0.18,
      buttons = list(

        list(
          label = "Pause",
          method = "animate",
          args = list(NULL, list(frame = list(duration = 0,
                                              redraw = TRUE),
                                 mode = "immediate"))
        )
      )
    )
  )
)
interactive_chart
```

*Figure 9 is an animated chart that tracks the cost of each coffee drink I bought or made and the happiness level I felt after receiving them. The visualization tracks the changes of the cost and happiness across the 7 weeks I took data*

It seems like money and happiness do go hand and hand for me when it comes to coffee. Excluding week 4 of my data collection, the scatter plot at each week has a positive correlation, indicating that an increase in the cost of the coffee is correlated with me being happier. Week 4 is an outlier in my data collection for this trend because I had multiple exams that week. Therefore, for most of the week, I was unbelievably stressed and that could have been a factor that contributed to the negative trend for that week alone. 

##### Figure 10 - Maybe I Need a Financial Advisor : 

```{r, echo=FALSE}
ggplot(coffee_me, aes(x=date., y=cost)) + geom_point() +
  geom_hline(yintercept=mean(coffee_me$cost), linetype="dashed", color = "darkred") + 
  geom_hline(yintercept=max(coffee_me$cost), linetype="dashed", color="navyblue") + 
  geom_hline(yintercept=3.01, linetype="dashed", color="darkgreen") +
  geom_text(x=as.Date("2023-10-25"), y=3.2, label="Average Amount of Money Adults in US Spends on Coffee Each Day",
            family="Courier", fontface="italic", color="darkgreen") + 
  geom_text(x=as.Date("2023-10-25"), y=4.3, label="Average Amount of Money I Spent on Coffee Each Day",
            family="Courier", fontface="italic", color="darkred") +
  geom_text(x=as.Date("2023-10-25"), y=7.9, label="Max Amount of Money I Spent on Coffee",
            family="Courier", fontface="italic", color="darkblue") +
  xlab("Date") + ylab("Cost (in Dollars)") + 
  ggtitle("Tracking the Cost of Each Drink Across October and November") +
  scale_y_continuous(breaks=seq(0,8,0.5))
```

*Figure 10 is a scatter plot that tracks the cost of each coffee drink across the 2 month period I tracked my coffee consumption habits. In addition, there are lines that indicate the average amount of money I spend on coffee, the average amount of money a coffee drinker in the US spends on coffee, and the highest amount of money I spent on coffee in these 2 months*

Looking at the amount of coffee I buy, I tracked the amount I spent across the months of October and November. Most of my coffee was bought from national chains or local cafes, and I rarely made coffee at home (the coffees made at home were considered to be a cost of 0 dollars). Most of my coffee prices range from 5.00 dollars to  6.50 dollars. The max amount of money I ever spent on coffee in these 2 months was 7.75 dollars. On average, I spend approximately 4.50 dollars on coffee every day. However, the average American does not spend this much money; instead, they only spend about $3.00 on coffee every day. The vast differences in the average money spent might indicate that I need to be a bit more responsible when it comes to my spending regarding my coffee because I am definitely out-spending the average American coffee drinker.

# Conclusion:  

In all, I was really shocked to see how I compare with the average coffee drinker in the United States. While I was under the impression that I was consuming significant amounts of coffee and caffeine, I do not come close to the average coffee drinker in the US (as indicated by figures 5 and 6). Similarly, my reasons for drinking coffee, my favorite drink, and where I normally drink coffee are also vastly different from the population of coffee drinkers in the US (as demonstrated in figures 1, 2, 3, 4, 7, 8) . In general, I am not considered a “normal” coffee drinker, rather I deviate from most of the norms of the average coffee drinker in the US. Furthermore, I proceeded to look into the costs and happiness I felt each time I drank coffee. I saw that as I spent more money on a coffee drink, my happiness also increased. However, it unfortunately will be irresponsible of me to continue increasing my spending on coffee as I already spend approximately 1.50 dollars more on coffee every day than the average coffee drinker in the United States. Part of the cause of the increase is due to my frequent purchases of coffee when I could potentially make my coffee at home or get it from the dining hall. 









